---
title: "part1.2reorganised"
author: "Jaey Tay"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r libraries}
# Load Libraries
library(tidyverse)
library(ggplot2)
library(viridis)
library(caret)
library(MASS)
library(randomForest)
library(pROC)
library(xtable)
library(glmnet)
library(gridExtra)
library(corrplot)
library(reshape2)
library(scales)


# Fix namespace conflicts
select <- dplyr::select
filter <- dplyr::filter
```

# Task 1.1

```{r data import}
# Set seed for reproducibility
set.seed(1)

# Import data
data <- read.csv("data-1.csv.gz")
cat("Data loaded:", nrow(data), "rows x", ncol(data), "columns\n")
```


```{r data-structure}
# Check structure
str(data, list.len = 10)
capture.output(str(data, list.len = 10))

# View first few rows
head(data[, 1:10])

missing_counts <- colSums(is.na(data))
cat("Total NA values:", sum(missing_counts))
```
The dataset has 215604 observations, and 220 spectral bands, which differs from the project description of 218680 observations and 218 spectral bands. The observations correspond to the reflective index of each particular band. No NA values are observed. 

```{r summary-stats}
# Class distribution
cat("Class distribution:")
class_dist <- table(data$land_type)
print(class_dist)

# Class proportions
cat("Class proportions:")
prop.table(class_dist) %>% round(3)
capture.output(prop.table(class_dist) %>% round(3))

library(knitr)

# Create a data frame with counts and proportions
class_summary <- data.frame(
  Class = names(class_dist),
  Count = as.vector(class_dist),
  Proportion = round(prop.table(class_dist), 3)
)

kable(class_summary, format = "latex", booktabs = TRUE)
```

```{r class-distribution, fig.height=5}
# Bar plot of class distribution
ggplot(data, aes(x = reorder(land_type, land_type, function(x) -length(x)))) +
  geom_bar(fill = "steelblue", alpha = 0.8) +
  geom_text(stat = 'count', aes(label = after_stat(count)), 
            vjust = -0.5, size = 3.5) +
  labs(title = "Distribution of Land Types",
       x = "Land Type",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold"))

# Check for class imbalance
cat("Class imbalance ratio (max/min):", 
    max(class_dist) / min(class_dist))
```
There are 8 distinct land types. The most prevalent land type are alpine meadows, making up 23.5% of the land area. On the other hand, dark rock is the least common land type, making up 4.70% of the land area. This would lead, to a certain extent, to some class imbalance, with there being a class imbalance factor of 5. We also observe that the proportion of snow/ice is 5.3%. This may present a problem for the subsequent glacier prediction task (1.4) as it becomes a very imbalanced binary problem. 

```{r spatial-distribution, fig.height=8}
# Create spatial visualization of land types
ggplot(data, aes(x = p_x, y = p_y, color = land_type)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "Spatial Distribution of Land Types",
       x = "X Coordinate",
       y = "Y Coordinate",
       color = "Land Type") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "right",
        legend.text = element_text(size = 8),
        legend.title = element_text(face = "bold"),
        legend.key.size = unit(1.2, "lines")) +
  coord_equal() +
  guides(color = guide_legend(override.aes = list(size = 5, alpha = 1)))
```
Here, we plot a map of our dataset, based on the x and y coordinates which were given. This serves as a useful cursory check for our data points. When observed beside the provided map of the area, we see that our map resembles the actual image of the region. 


```{r glacier-binary, fig.height=8, fig.width=10}
# Highlight glacier/snow regions for T1.4
data_glacier <- data %>%
  mutate(is_glacier = ifelse(land_type == "snow / ice", "Glacier/Snow", "Other"))

ggplot(data_glacier, aes(x = p_x, y = p_y, color = is_glacier)) +
  geom_point(size = 0.3, alpha = 0.6) +
  scale_color_manual(values = c("Glacier/Snow" = "#00D9FF", "Other" = "gray70")) +
  scale_y_reverse() +
  labs(title = "Glacier Ice Detection Task (T1.4): Binary Classification",
       subtitle = "Snow/ice (5.3%) vs. all other land types (94.7%)",
       x = "X Coordinate (pixels)",
       y = "Y Coordinate (pixels)",
       color = "Class") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 10),
        legend.position = "right",
        legend.text = element_text(size = 10),
        legend.title = element_text(face = "bold")) +
  coord_equal() +
  guides(color = guide_legend(override.aes = list(size = 5, alpha = 1)))
```
This visualisation highlights the sparsity of snow/ice land types. The class imbalance (5.3% positive class) suggests that evaluation metrics such as F1-score(?) or balanced accuracy will be more appropriate than overall accuracy, as a naive classifier predicting "non-glacier" for all pixels would achieve 94.7% accuracy while being completely uninformative.


```{r spectral-signatures, fig.height=6}

band_cols <- grep("^Band_", names(data), value = TRUE)
spectral_data <- data[, band_cols]
# Calculate mean spectral signature for each class
spectral_means <- data %>%
  dplyr::select(land_type, all_of(band_cols)) %>%
  group_by(land_type) %>%
  summarise(across(everything(), mean)) %>%
  pivot_longer(cols = -land_type, names_to = "Band", values_to = "Reflectance") %>%
  mutate(Band_num = as.numeric(gsub("Band_", "", Band)))

# Plot mean spectral signatures
ggplot(spectral_means, aes(x = Band_num, y = Reflectance, color = land_type)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "Mean Spectral Signatures by Land Type",
       subtitle = "Average reflectance across wavelength bands (≈ 420-2450 nm)",
       x = "Band Number",
       y = "Mean Reflectance",
       color = "Land Type") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

The mean spectral signatures reveal clear separability between certain land types. Snow and ice exhibit higher reflectance on the lower bands of the spectrum. Land types which feature vegetation (alpine meadow, alpine tundra, valley floor and veg-scree mix) display similar reflective patterns, suggesting these may be more challenging to discriminate. On the other hand, the different kinds of rock exhibit more distinct spectral profiles, suggesting that their reflective features should provide stronger discriminative power for classification.


```{r spectral-variability, fig.height=6}
# Calculate standard deviation for each class
spectral_sd <- data %>%
  dplyr::select(land_type, all_of(band_cols)) %>%
  group_by(land_type) %>%
  summarise(across(everything(), sd)) %>%
  pivot_longer(cols = -land_type, names_to = "Band", values_to = "SD") %>%
  mutate(Band_num = as.numeric(gsub("Band_", "", Band)))

# Plot spectral variability
ggplot(spectral_sd, aes(x = Band_num, y = SD, color = land_type)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "Spectral Variability by Land Type",
       subtitle = "Standard deviation of reflectance across wavelength bands",
       x = "Band Number",
       y = "Standard Deviation",
       color = "Land Type") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```
For snow/ice land types, we observe that the standard deviation of reflectivity increases a lot as we approach the lower bands of the spectrum. At lower bands, dark rock and scree/sunlit rock also exhibit higher standard deviation of reflectivity than other land masses. As we get to higher bands, we see that the standard deviation of reflectivity generally remains relatively lower for all land types. 





# Task 1.2

```{r data preparation}
# Clean factor levels for caret compatibility
data$land_type <- as.factor(data$land_type)
current_levels <- levels(data$land_type)
temp_levels <- gsub("/", ".", current_levels, fixed = TRUE)
temp_levels <- gsub(" ", ".", temp_levels, fixed = TRUE)
cleaned_levels <- make.names(temp_levels, unique = TRUE)
levels(data$land_type) <- cleaned_levels

# Create spatial blocks for train/test split
nx <- 4
ny <- 4

data <- data %>%
  mutate(
    x_block = cut(p_x, breaks = nx, labels = FALSE),
    y_block = cut(p_y, breaks = ny, labels = FALSE),
    block_id = paste(x_block, y_block, sep = "_")
  )

# Split into training and test blocks
unique_blocks <- unique(data$block_id)
set.seed(1)
test_blocks <- sample(unique_blocks, 3)
remaining_blocks <- setdiff(unique_blocks, test_blocks)
train_blocks <- sample(remaining_blocks, 9)

train_data <- data %>% filter(block_id %in% train_blocks)
test_data <- data %>% filter(block_id %in% test_blocks)

cat("\nData Split:\n")
cat("  Training pixels:", nrow(train_data), "\n")
cat("  Test pixels:", nrow(test_data), "\n")

# Define CV control (used by all models)
K_FOLDS <- 5
set.seed(123)
folds_index <- createFolds(
  y = train_data$land_type,
  k = K_FOLDS,
  list = TRUE,
  returnTrain = TRUE
)

ctrl <- trainControl(
  method = "cv",
  number = K_FOLDS,
  index = folds_index,
  savePredictions = FALSE,
  classProbs = TRUE,
  verboseIter = FALSE
)

# Extract predictor columns
pred_cols <- grep("^Band", colnames(train_data), value = TRUE)
```

```{r model evaluation function}
evaluate_model <- function(model, test_data, model_name, 
                          save_latex = FALSE, file_prefix = NULL) {
  
  # Predictions
  preds <- predict(model, test_data)
  probs <- predict(model, test_data, type = "prob")
  
  # Confusion Matrix
  cm <- confusionMatrix(preds, test_data$land_type)
  
  # One-vs-All Metrics
  classes <- levels(test_data$land_type)
  ova <- data.frame(
    Class = classes,
    Balanced_Acc = numeric(length(classes)),
    AUC = numeric(length(classes)),
    F1 = numeric(length(classes))
  )
  
  for (i in seq_along(classes)) {
    k <- classes[i]
    binary_truth <- factor(test_data$land_type == k, levels = c(TRUE, FALSE))
    binary_prob <- probs[, k]
    binary_pred <- factor(preds == k, levels = c(TRUE, FALSE))
    
    cm_ova <- confusionMatrix(binary_pred, binary_truth, positive = "TRUE")
    ova$Balanced_Acc[i] <- cm_ova$byClass["Balanced Accuracy"]
    ova$F1[i] <- cm_ova$byClass["F1"]
    
    roc_obj <- roc(binary_truth, binary_prob, levels = c(FALSE, TRUE), quiet = TRUE)
    ova$AUC[i] <- as.numeric(auc(roc_obj))
  }
  
  # Calculate averages
  avg_bal_acc <- mean(ova$Balanced_Acc, na.rm = TRUE)
  avg_auc <- mean(ova$AUC, na.rm = TRUE)
  avg_f1 <- mean(ova$F1, na.rm = TRUE)
  
  # Console output
  cat("\n", rep("=", 60), "\n", sep = "")
  cat(model_name, "\n")
  cat(rep("=", 60), "\n\n", sep = "")
  
  cat("Confusion Matrix:\n")
  print(cm$table)
  cat("\n")
  
  cat(sprintf("Overall Accuracy:           %.4f\n", cm$overall["Accuracy"]))
  cat(sprintf("Misclassification Error:    %.4f\n", 1 - cm$overall["Accuracy"]))
  cat(sprintf("Avg. Balanced Accuracy:     %.4f\n", avg_bal_acc))
  cat(sprintf("Avg. AUC:                   %.4f\n", avg_auc))
  cat(sprintf("Avg. F1 Score:              %.4f\n", avg_f1))
  cat("\n")
  
  # LaTeX output
  if (save_latex && !is.null(file_prefix)) {
    summary_metrics <- data.frame(
      Metric = c("Overall Accuracy", "Misclassification Error", "Avg. Balanced Accuracy", "Avg. AUC", "Avg. F1 Score"),
      Value = round(c(cm$overall["Accuracy"], 
                     1 - cm$overall["Accuracy"],
                     avg_bal_acc, avg_auc, avg_f1), 4)
    )
    
    cat("\n% LaTeX Table for", model_name, "\n")
    cat(capture.output(print(xtable(summary_metrics,
                                    caption = paste(model_name, "Performance"),
                                    label = paste0("tab:", file_prefix)),
                            include.rownames = FALSE,
                            booktabs = TRUE)), sep = "\n")
  }
  
  # Return results
  invisible(list(
    cm = cm,
    ova = ova,
    avg_bal_acc = avg_bal_acc,
    avg_auc = avg_auc,
    avg_f1 = avg_f1
  ))
}
```


```{r model training}
# Storage for all results
all_results <- list()

# Extract predictor columns
pred_cols <- grep("^Band", colnames(train_data), value = TRUE)
```


```{r LDA model training}
# LDA - Original Features
train_lda_orig <- train_data[, c(pred_cols, "land_type")]

cat("\nTraining LDA - Original Features...\n")
model_lda_orig <- train(
  land_type ~ .,
  data = train_lda_orig,
  method = "lda",
  trControl = ctrl,
  metric = "Accuracy"
)
all_results$lda_orig <- evaluate_model(
  model_lda_orig, test_data,
  "LDA - Original", TRUE, "lda_orig"
)

# LDA - PCA Features
train_lda_pca <- train_data[, c(pred_cols, "land_type")]

cat("\nTraining LDA - PCA Features...\n")
model_lda_pca <- train(
  land_type ~ .,
  data = train_lda_pca,
  method = "lda",
  preProcess = c("center", "scale", "pca"),
  preProcOptions = list(pcaComp = 10),
  trControl = ctrl,
  metric = "Accuracy"
)
all_results$lda_pca <- evaluate_model(
  model_lda_pca, test_data,
  "LDA - PCA", TRUE, "lda_pca"
)
```


```{r logistic model training}
train_log_orig <- train_data[, c(pred_cols, "land_type")]

ridge_grid_orig <- expand.grid(
  alpha = 0,
  lambda = seq(0.001, 1, length = 20)
)

cat("\nTraining Logistic Regression - Original Features...\n")

model_log_orig <- train(
  land_type ~ .,
  data = train_log_orig,
  method = "glmnet",
  family = "multinomial",
  tuneGrid = ridge_grid_orig,
  trControl = ctrl,
  metric = "Accuracy"
)

all_results$log_orig <- evaluate_model(
  model_log_orig, test_data,
  "Logistic - Original", TRUE, "log_orig"
)


# Logistic - PCA Features
cat("\nTraining Logistic Regression - PCA Features...\n")

train_log_pca <- train_data[, c(pred_cols, "land_type")]

ridge_grid_pca <- expand.grid(
  alpha = 0,
  lambda = seq(0.001, 0.5, length = 20)
)

model_log_pca <- train(
  land_type ~ .,
  data = train_log_pca,
  method = "multinom",
  trace = FALSE,
  trControl = ctrl,                
  metric = "Accuracy",
  preProcess = c("center", "scale", "pca"),
  preProcOptions = list(pcaComp = 10)
)

all_results$log_pca <- evaluate_model(
  model_log_pca, test_data,
  "Logistic - PCA", TRUE, "log_pca"
)
```


```{r QDA model training}

# QDA - Original Features
train_qda_orig <- train_data[, c(pred_cols, "land_type")]

cat("\nTraining QDA - Original Features...\n")
model_qda_orig <- train(
  land_type ~ .,
  data = train_qda_orig,
  method = "qda",
  trControl = ctrl,
  metric = "Accuracy"
)
all_results$qda_orig <- evaluate_model(
  model_qda_orig, test_data,
  "QDA - Original", TRUE, "qda_orig"
)

# QDA - PCA Features
train_qda_pca <- train_data[, c(pred_cols, "land_type")]

cat("\nTraining QDA - PCA Features...\n")
model_qda_pca <- train(
  land_type ~ .,
  data = train_qda_pca,
  method = "qda",
  preProcess = c("center", "scale", "pca"),
  preProcOptions = list(pcaComp = 10),
  trControl = ctrl,
  metric = "Accuracy"
)
all_results$qda_pca <- evaluate_model(
  model_qda_pca, test_data,
  "QDA - PCA", TRUE, "qda_pca"
)
```


```{r knn model training}
# k-NN - Original Features (10% sample)
cat("\nTraining k-NN - Original Features (10% sample)...\n")

set.seed(0)
knn_sample_idx <- createDataPartition(
  train_data$land_type,
  p = 0.10,
  list = FALSE
)

train_knn_orig <- train_data[knn_sample_idx, c(pred_cols, "land_type")]

cat("k-NN Training Sample Size:", nrow(train_knn_orig), "\n")
cat("Original Training Size:", nrow(train_data), "\n")

# Create NEW folds for the sampled data
knn_folds_orig <- createFolds(
  y = train_knn_orig$land_type,
  k = K_FOLDS,
  list = TRUE,
  returnTrain = TRUE
)

ctrl_knn_orig <- trainControl(
  method = "cv",
  number = K_FOLDS,
  index = knn_folds_orig,  # Use folds from sampled data!
  savePredictions = FALSE,
  classProbs = TRUE,
  verboseIter = FALSE
)

model_knn_orig <- train(
  land_type ~ .,
  data = train_knn_orig,
  method = "knn",
  tuneGrid = data.frame(k = c(3, 5, 7, 11, 15)),
  trControl = ctrl_knn_orig,  # Use the sampled-data ctrl
  metric = "Accuracy"
)

cat("\nk-NN (Original) Training Complete.\n")

all_results$knn_orig <- evaluate_model(
  model_knn_orig, test_data,
  "k-NN - Original (10% sample)", TRUE, "knn_orig"
)

# k-NN - PCA Features (full data)
cat("\nTraining k-NN - PCA Features (full data)...\n")

# Manually do PCA preprocessing
pca_prep_knn <- preProcess(
  train_data[, pred_cols],
  method = c("center", "scale", "pca"),
  pcaComp = 10
)

# Apply PCA to training data
train_knn_pca_transformed <- predict(pca_prep_knn, train_data[, pred_cols])
train_knn_pca <- data.frame(train_knn_pca_transformed, land_type = train_data$land_type)

# Train k-NN on PCA-transformed data
model_knn_pca <- train(
  land_type ~ .,
  data = train_knn_pca,
  method = "knn",
  tuneGrid = data.frame(k = c(3, 5, 7, 11, 15, 21)),
  trControl = ctrl,
  metric = "Accuracy"
)

cat("\nk-NN (PCA) Training Complete.\n")

test_knn_pca_transformed <- predict(pca_prep_knn, test_data[, pred_cols])
test_knn_pca <- data.frame(test_knn_pca_transformed, land_type = test_data$land_type)

all_results$knn_pca <- evaluate_model(
  model_knn_pca, 
  test_knn_pca,  # Use transformed test data!
  "k-NN - PCA", TRUE, "knn_pca")
```



```{r gbdt model training}
# GBDT (XGBoost) - Original Features (10% sample)

cat("\nTraining GBDT - Original Features (10% sample)...\n")

set.seed(0)
gbdt_sample_idx <- createDataPartition(train_data$land_type, p = 0.10, list = FALSE)
train_gbdt_orig <- train_data[gbdt_sample_idx, c(pred_cols, "land_type")]

cat("GBDT Original Training Sample Size:", nrow(train_gbdt_orig), "\n")

gbdt_folds_orig <- createFolds(
  y = train_gbdt_orig$land_type,
  k = K_FOLDS,
  list = TRUE,
  returnTrain = TRUE
)

ctrl_gbdt_orig <- trainControl(
  method = "cv",
  number = K_FOLDS,
  index = gbdt_folds_orig,
  savePredictions = FALSE,
  classProbs = TRUE,
  verboseIter = TRUE
)

# XGBoost tuning grid
gbdt_grid <- expand.grid(
  nrounds = c(100, 200),           # Number of boosting iterations
  max_depth = c(3, 6, 9),          # Tree depth
  eta = c(0.1, 0.3),               # Learning rate
  gamma = 0,                        # Minimum loss reduction
  colsample_bytree = 0.8,          # Subsample ratio of columns
  min_child_weight = 1,            # Minimum sum of instance weight
  subsample = 0.8                  # Subsample ratio of training instances
)

set.seed(0)
model_gbdt_orig <- train(
  land_type ~ .,
  data = train_gbdt_orig,
  method = "xgbTree",
  trControl = ctrl_gbdt_orig,
  tuneGrid = gbdt_grid,
  metric = "Accuracy",
  verbosity = 0  # Suppress XGBoost warnings
)

cat("\nGBDT (Original) Training Complete.\n")
cat("Best parameters:\n")
print(model_gbdt_orig$bestTune)

all_results$gbdt_orig <- evaluate_model(
  model_gbdt_orig, test_data,
  "GBDT - Original (10% sample)", TRUE, "gbdt_orig"
)


# GBDT (XGBoost) - PCA Features (full data)

cat("\nTraining GBDT - PCA Features (full data)...\n")

train_gbdt_pca <- train_data[, c(pred_cols, "land_type")]

cat("GBDT PCA Training Size:", nrow(train_gbdt_pca), "\n")

gbdt_folds_pca <- createFolds(
  y = train_gbdt_pca$land_type,
  k = K_FOLDS,
  list = TRUE,
  returnTrain = TRUE
)

ctrl_gbdt_pca <- trainControl(
  method = "cv",
  number = K_FOLDS,
  index = gbdt_folds_pca,
  savePredictions = FALSE,
  classProbs = TRUE,
  verboseIter = TRUE
)

# Smaller grid for PCA (fewer features = faster)
gbdt_grid_pca <- expand.grid(
  nrounds = c(100, 200, 300),
  max_depth = c(3, 6, 9),
  eta = c(0.1, 0.3),
  gamma = 0,
  colsample_bytree = 0.8,
  min_child_weight = 1,
  subsample = 0.8
)

set.seed(0)
model_gbdt_pca <- train(
  land_type ~ .,
  data = train_gbdt_pca,
  method = "xgbTree",
  preProcess = c("center", "scale", "pca"),
  preProcOptions = list(pcaComp = 10),
  trControl = ctrl_gbdt_pca,
  tuneGrid = gbdt_grid_pca,
  metric = "Accuracy",
  verbosity = 0
)

cat("\nGBDT (PCA) Training Complete.\n")
cat("Best parameters:\n")
print(model_gbdt_pca$bestTune)

all_results$gbdt_pca <- evaluate_model(
  model_gbdt_pca, test_data,
  "GBDT - PCA", TRUE, "gbdt_pca"
)
```


```{r RF model training}
# Random Forest - Original Features (10% sample)
cat("\nTraining Random Forest - Original Features (10% sample)...\n")

set.seed(0)
rf_sample_idx <- createDataPartition(
  train_data$land_type,
  p = 0.10,
  list = FALSE
)

train_rf_orig <- train_data[rf_sample_idx, c(pred_cols, "land_type")]

cat("RF Training Sample Size:", nrow(train_rf_orig), "\n")
cat("Original Training Size:", nrow(train_data), "\n")

# Create NEW folds for the sampled data
rf_folds_orig <- createFolds(
  y = train_rf_orig$land_type,
  k = K_FOLDS,
  list = TRUE,
  returnTrain = TRUE
)

ctrl_rf_orig <- trainControl(
  method = "cv",
  number = K_FOLDS,
  index = rf_folds_orig,
  savePredictions = FALSE,
  classProbs = TRUE,
  verboseIter = TRUE
)

set.seed(0)
model_rf_orig <- train(
  land_type ~ .,
  data = train_rf_orig,
  method = "rf",
  tuneGrid = expand.grid(mtry = c(10, 15, 20, 30, 50)),
  trControl = ctrl_rf_orig,
  ntree = 500,
  importance = TRUE,
  metric = "Accuracy"
)

cat("\nRandom Forest (Original) Training Complete.\n")

all_results$rf_orig <- evaluate_model(
  model_rf_orig, test_data,
  "Random Forest - Original (10% sample)", TRUE, "rf_orig"
)


# Random Forest - PCA Features (full data)
cat("\nTraining Random Forest - PCA Features (full data)...\n")

train_rf_pca <- train_data[, c(pred_cols, "land_type")]

cat("RF PCA Training Size:", nrow(train_rf_pca), "\n")

# Create folds for full data
rf_folds_pca <- createFolds(
  y = train_rf_pca$land_type,
  k = K_FOLDS,
  list = TRUE,
  returnTrain = TRUE
)

ctrl_rf_pca <- trainControl(
  method = "cv",
  number = K_FOLDS,
  index = rf_folds_pca,
  savePredictions = FALSE,
  classProbs = TRUE,
  verboseIter = TRUE
)

set.seed(0)
model_rf_pca <- train(
  land_type ~ .,
  data = train_rf_pca,
  method = "rf",
  preProcess = c("center", "scale", "pca"),
  preProcOptions = list(pcaComp = 10),
  tuneGrid = expand.grid(mtry = c(2, 3, 5, 7)),
  trControl = ctrl_rf_pca,
  ntree = 500,
  importance = TRUE,
  metric = "Accuracy"
)

cat("\nRandom Forest (PCA) Training Complete.\n")

all_results$rf_pca <- evaluate_model(
  model_rf_pca, test_data,
  "Random Forest - PCA", TRUE, "rf_pca"
)
```


```{r SVM model training}
# SVM - Original Features (10% sample)
cat("\nTraining SVM - Original Features (10% sample)...\n")
set.seed(0)
svm_sample_idx <- createDataPartition(train_data$land_type, p = 0.10, list = FALSE)
train_svm_orig <- train_data[svm_sample_idx, c(pred_cols, "land_type")]
cat("SVM Original Training Sample Size:", nrow(train_svm_orig), "\n")
svm_folds_orig <- createFolds(
  y = train_svm_orig$land_type,
  k = K_FOLDS,
  list = TRUE,
  returnTrain = TRUE
)
ctrl_svm_orig <- trainControl(
  method = "cv",
  number = K_FOLDS,
  index = svm_folds_orig,
  savePredictions = FALSE,
  classProbs = TRUE,
  verboseIter = TRUE
)
model_svm_orig <- train(
  land_type ~ .,
  data = train_svm_orig,
  method = "svmRadial",
  tuneGrid = expand.grid(C = c(0.1, 1, 10), sigma = c(0.01, 0.1, 1)),
  trControl = ctrl_svm_orig,
  metric = "Accuracy"
)
cat("\nSVM (Original) Training Complete.\n")
all_results$svm_orig <- evaluate_model(
  model_svm_orig, test_data,
  "SVM - Original (10% sample)", TRUE, "svm_orig"
)
# SVM - PCA Features (10% sample)
cat("\nTraining SVM - PCA Features (10% sample)...\n")
set.seed(0)
svm_sample_idx_pca <- createDataPartition(train_data$land_type, p = 0.10, list = FALSE)
train_svm_pca <- train_data[svm_sample_idx_pca, c(pred_cols, "land_type")]
cat("SVM PCA Training Sample Size:", nrow(train_svm_pca), "\n")
svm_folds_pca <- createFolds(
  y = train_svm_pca$land_type,
  k = K_FOLDS,
  list = TRUE,
  returnTrain = TRUE
)
ctrl_svm_pca <- trainControl(
  method = "cv",
  number = K_FOLDS,
  index = svm_folds_pca,
  savePredictions = FALSE,
  classProbs = TRUE,
  verboseIter = TRUE
)
model_svm_pca <- train(
  land_type ~ .,
  data = train_svm_pca,
  method = "svmRadial",
  preProcess = c("center", "scale", "pca"),
  preProcOptions = list(pcaComp = 10),
  tuneGrid = expand.grid(C = c(0.1, 1, 10, 100), sigma = c(0.01, 0.1, 1)),
  trControl = ctrl_svm_pca,
  metric = "Accuracy"
)
cat("\nSVM (PCA) Training Complete.\n")
all_results$svm_pca <- evaluate_model(
  model_svm_pca, test_data,
  "SVM - PCA (10% sample)", TRUE, "svm_pca"
)
```

```{r summary table}
# Build summary table from all_results
build_summary_table <- function(all_results) {
  
  # Define model order and display names
  models <- list(
    lda_orig = "LDA (Original)",
    lda_pca = "LDA (PCA)",
    log_orig = "Logistic (Original)",
    log_pca = "Logistic (PCA)",
    qda_orig = "QDA (Original)",
    qda_pca = "QDA (PCA)",
    knn_orig = "$k$-NN (Original)",
    knn_pca = "$k$-NN (PCA)",
    gbdt_orig = "xGBoost (Original)",
    gbdt_pca = "xGBoost (PCA)",
    rf_orig = "RF (Original)",
    rf_pca = "RF (PCA)",
    svm_orig = "SVM (Original)",
    svm_pca = "SVM (PCA)"
  )
  
  # Extract metrics
  summary_df <- data.frame(
    Model = character(),
    Accuracy = numeric(),
    Error_Rate = numeric(),
    Bal_Acc = numeric(),
    AUC = numeric(),
    F1 = numeric(),
    stringsAsFactors = FALSE
  )
  
  for (key in names(models)) {
    if (key %in% names(all_results)) {
      res <- all_results[[key]]
      acc <- res$cm$overall["Accuracy"]
      summary_df <- rbind(summary_df, data.frame(
        Model = models[[key]],
        Accuracy = round(acc, 3),
        Error_Rate = round(1 - acc, 3),
        Bal_Acc = round(res$avg_bal_acc, 3),
        AUC = round(res$avg_auc, 3),
        F1 = round(res$avg_f1, 3),
        stringsAsFactors = FALSE
      ))
    }
  }
  
  rownames(summary_df) <- NULL
  return(summary_df)
}

# Generate LaTeX table with bold for best values
generate_latex_table <- function(summary_df) {
  
  # Find best for each metric (max for most, min for error rate)
  best_acc <- max(summary_df$Accuracy)
  best_err <- min(summary_df$Error_Rate)  # Lower is better!
  best_bal <- max(summary_df$Bal_Acc)
  best_auc <- max(summary_df$AUC)
  best_f1 <- max(summary_df$F1)
  
  # Format function: bold if best
  fmt <- function(val, best) {
    if (val == best) {
      sprintf("\\textbf{%.3f}", val)
    } else {
      sprintf("%.3f", val)
    }
  }
  
  # Build LaTeX
  cat("\\begin{table}[htbp]\n")
  cat("\\centering\n")
  cat("\\caption{Summary of evaluation results}\n")
  cat("\\label{tab:all_results}\n")
  cat("\\begin{tabular}{l|ccccc}\n")
  cat("\\toprule\n")
  cat("Model & Accuracy & Error Rate & Bal. Acc & AUC & F1 \\\\\n")
  cat("\\midrule\n")
  
  for (i in 1:nrow(summary_df)) {
    row <- summary_df[i, ]
    cat(sprintf("%s & %s & %s & %s & %s & %s \\\\\n",
                row$Model,
                fmt(row$Accuracy, best_acc),
                fmt(row$Error_Rate, best_err),
                fmt(row$Bal_Acc, best_bal),
                fmt(row$AUC, best_auc),
                fmt(row$F1, best_f1)))
  }
  
  cat("\\bottomrule\n")
  cat("\\end{tabular}\n")
  cat("\\end{table}\n")
}

# Run it
summary_table <- build_summary_table(all_results)
print(summary_table)

cat("\n\n% === LATEX TABLE (copy below) ===\n\n")
generate_latex_table(summary_table)
```

```{r mypredict function}

# Uses existing objects:
# - pca_prep_knn: preProcess object for PCA transformation
# - model_knn_pca: trained k-NN model

mypredict <- function(newdata) {
  
  # Clean factor levels for caret compatibility
  newdata$land_type <- as.factor(newdata$land_type)
  current_levels <- levels(newdata$land_type)
  temp_levels <- gsub("/", ".", current_levels, fixed = TRUE)
  temp_levels <- gsub(" ", ".", temp_levels, fixed = TRUE)
  cleaned_levels <- make.names(temp_levels, unique = TRUE)
  levels(newdata$land_type) <- cleaned_levels
  
  # Get predictor columns
  pred_cols <- grep("^Band", colnames(newdata), value = TRUE)
  
  # Apply same PCA transformation used in training
  newdata_pca <- predict(pca_prep_knn, newdata[, pred_cols])
  
  # Predict using trained k-NN model
  predictions <- predict(model_knn_pca, newdata_pca)
  
  return(predictions)
}


test_predictions <- mypredict(test_data)
test_accuracy <- mean(test_predictions == test_data$land_type)
cat(sprintf("Test accuracy: %.4f\n", test_accuracy))
```


# Task 1.3

```{r performance visualisations}
# Get best model key and clean the name
best_key <- names(all_results)[which.max(sapply(all_results, function(x) x$cm$overall["Accuracy"]))]
best_model_name <- summary_table$Model[which.max(summary_table$Accuracy)]

best_model_name_clean <- gsub("\\$", "", best_model_name)  # Remove $ signs

# Per-Class Performance (Best Model) 
best_ova <- all_results[[best_key]]$ova

ova_long <- best_ova %>%
  pivot_longer(cols = c(Balanced_Acc, AUC, F1),
               names_to = "Metric", values_to = "Value")

p1 <- ggplot(ova_long, aes(x = Class, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.8) +
  geom_text(aes(label = sprintf("%.2f", Value)), 
            position = position_dodge(width = 0.9),
            vjust = -0.5, 
            size = 2.5) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +  # Add space for labels
  labs(title = paste("Per-Class Performance:", best_model_name_clean),
       x = "Land Type", y = "Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

print(p1)

#PCA Scatter Plot (First 2 PCs, colored by class)

pca_df <- data.frame(
  train_knn_pca[, 1:3],  # PC1, PC2, PC3 already computed
  land_type = train_knn_pca$land_type
)
colnames(pca_df)[1:3] <- c("PC1", "PC2", "PC3")

pca_var <- prcomp(train_data[, pred_cols], center = TRUE, scale. = TRUE)
var_explained <- summary(pca_var)$importance

cat("Variance explained by PC1:", round(var_explained[2, 1] * 100, 1), "%\n")
cat("Variance explained by PC2:", round(var_explained[2, 2] * 100, 1), "%\n")
cat("Cumulative (PC1+PC2):", round(var_explained[3, 2] * 100, 1), "%\n")
cat("Cumulative (PC1-PC10):", round(var_explained[3, 10] * 100, 1), "%\n")



p2 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = land_type)) +
  geom_point(alpha = 0.3, size = 0.8) +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "PCA: First Two Principal Components",
       x = "PC1", y = "PC2", color = "Land Type") +
  theme_minimal() +
  theme(legend.position = "right") +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))

print(p2)
```

# Task 1.4


In glaciology and climate research, an important task is monitoring glacier extent over time. Using the provided hyperspectral satellite data, frame glacier-ice detection as a binary task with glacier ice = positive and all other land types = negative. Choose and justify an evaluation metric appropriate for this setting (e.g. F1 with glacier as the positive class). Train and evaluate three classifiers of your choice with the goal of improving your chosen performance metric. You may use any classifier or combination of methods covered in the course, including those listed in T1.2, as well as methods like bagging, boosting, and regularisation.

```{r}
# cleaning data for caret
dataset$land_type <- as.factor(dataset$land_type)

current_levels <- levels(dataset$land_type)

temp_levels <- gsub("/", ".", current_levels, fixed = TRUE)
temp_levels <- gsub(" ", ".", temp_levels, fixed = TRUE)
cleaned_levels <- make.names(temp_levels, unique = TRUE)

levels(dataset$land_type) <- cleaned_levels

# check
levels(dataset$land_type)
```
```{r}
dataset$glacier <- ifelse(dataset$land_type == "snow...ice", 1, 0)

# Convert to factor
dataset$glacier <- factor(dataset$glacier, levels = c(0,1))
table(dataset$land_type, dataset$glacier)
```

```{r}
# LDA
N <- 10  # number of PCs
set.seed(123)

# Create 5-fold stratified CV based on glacier column
folds_lda <- createFolds(dataset$glacier, k = 5, list = TRUE, returnTrain = TRUE)

# Initialize vector to store LDA F1 scores
f1_scores_lda <- numeric(length(folds_lda))

# Loop over folds
for(i in seq_along(folds_lda)) {
  
  # Training and test sets
  train_idx_lda <- folds_lda[[i]]
  train_data_lda <- dataset[train_idx_lda, ]
  test_data_lda  <- dataset[-train_idx_lda, ]
  
  # PCA on training fold
  pca_lda <- prcomp(train_data_lda[, band_cols], center = TRUE, scale. = TRUE)
  
  # Top N PCs
  train_pcs_lda <- as.data.frame(pca_lda$x[, 1:N])
  test_pcs_lda  <- as.data.frame(predict(pca_lda, newdata = test_data_lda[, band_cols])[, 1:N])
  
  # Add target variable
  train_pcs_lda$glacier_lda <- train_data_lda$glacier
  test_pcs_lda$glacier_lda  <- test_data_lda$glacier
  

  # Fit LDA model
  model_lda <- lda(glacier_lda ~ ., data = train_pcs_lda)
  
  # Predict class probabilities
  pred_prob_lda <- predict(model_lda, newdata = test_pcs_lda)$posterior[, "1"]
  
  # Convert to binary predictions
  pred_lda <- ifelse(pred_prob_lda > 0.5, 1, 0)
  pred_lda <- factor(pred_lda, levels = c(0,1))
  true_lda <- factor(test_pcs_lda$glacier_lda, levels = c(0,1))
  
  # Compute F1 score
  f1_scores_lda[i] <- F_meas(pred_lda, true_lda, relevant = "1")
  
  cat("Fold", i, "LDA F1:", round(f1_scores_lda[i], 6), "\n")
}

# Average F1 across folds
cat("Average LDA F1 across 5 folds:", round(mean(f1_scores_lda), 6), "\n")

```
Hyperspectral data often violates LDA assumptions because the spectral bands are not perfectly normal and the Covariance of glacier vs non-glacier pixels differs.

Logistic regression doesn’t assume normality, so it just models a linear boundary in feature space and will likely perform better than the LDA model.

```{r}
# logistic regression
# Store F1 for each fold

# Identify spectral bands
# Replace with your actual band column names
band_cols <- grep("^Band_", names(dataset), value = TRUE)

# Create 5-fold stratified CV
set.seed(123)
folds <- createFolds(dataset$glacier, k = 5, list = TRUE, returnTrain = TRUE)


# Initialize vector to store F1 scores
f1_scores <- numeric(length(folds))

# loops over folds
for(i in seq_along(folds)){
  
  # Split training and test sets
  train_idx <- folds[[i]]
  train_data <- dataset[train_idx, ]
  test_data  <- dataset[-train_idx, ]
  

  # PCA on training set only
  pca <- prcomp(train_data[, band_cols], center = TRUE, scale. = TRUE)
  
  # Project training & test data onto top 10 PCs
  N <- 10
  train_pcs <- as.data.frame(pca$x[, 1:N])
  test_pcs  <- as.data.frame(predict(pca, newdata = test_data[, band_cols])[, 1:N])
  
  # Add target variable
  train_pcs$glacier <- train_data$glacier
  test_pcs$glacier  <- test_data$glacier
  

  # Logistic regression
  model_logit <- glm(glacier ~ ., data = train_pcs, family = binomial)
  
  # Predict probabilities on test set
  prob <- predict(model_logit, newdata = test_pcs, type = "response")
  
  # Convert to binary predictions (threshold = 0.5)
  pred <- ifelse(prob > 0.5, 1, 0)
  pred <- factor(pred, levels = c(0,1))
  true <- factor(test_pcs$glacier, levels = c(0,1))
  

  # Compute F1 score
  f1_scores[i] <- F_meas(pred, true, relevant = "1")
  cat("Fold", i, "F1:", round(f1_scores[i], 6), "\n")
}


# Average F1 across folds
cat("Average F1 across 5 folds:", round(mean(f1_scores), 6), "\n")


```
Logistic regression already is performing really well with an average F1 score of 0.998. 

```{r}
# random forest
set.seed(123)
folds_rf <- createFolds(dataset$glacier, k = 5, list = TRUE, returnTrain = TRUE)

# Select spectral bands (example: Band_1 to Band_218)
band_cols_rf <- grep("^Band_", names(dataset), value = TRUE)

# Initialize vector to store F1 scores
f1_scores_rf <- numeric(length(folds_rf))

# Loop over folds
for(i in seq_along(folds_rf)) {
  
  # Training and test sets
  train_idx_rf <- folds_rf[[i]]
  train_data_rf <- dataset[train_idx_rf, ]
  test_data_rf  <- dataset[-train_idx_rf, ]

  # Fit Random Forest
  rf_model <- randomForest(
    x = train_data_rf[, band_cols_rf],
    y = train_data_rf$glacier,
    ntree = 500,          # number of trees
    mtry = floor(sqrt(length(band_cols_rf))),  # default: sqrt(num features)
    importance = TRUE
  )
  
  # Predict on test set
  pred_rf <- predict(rf_model, newdata = test_data_rf[, band_cols_rf])
  
  # Convert to factor with correct levels
  pred_rf <- factor(pred_rf, levels = c(0,1))
  true_rf <- factor(test_data_rf$glacier, levels = c(0,1))
  
  # Compute F1 score
  f1_scores_rf[i] <- F_meas(pred_rf, true_rf, relevant = "1")
  
  cat("Fold", i, "RF F1:", round(f1_scores_rf[i], 3), "\n")
}


# Average F1 across folds
cat("Average RF F1 across 5 folds:", round(mean(f1_scores_rf), 3), "\n")

#  Feature importance
importance(rf_model)
varImpPlot(rf_model)
```
